\documentclass[11pt]{article}

% Change "review" to "final" to generate the final (sometimes called camera-ready) version.
% Change to "preprint" to generate a non-anonymous version with page numbers.
\usepackage[review]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{subcaption}

\title{Who Wrote This? Identity Bias in LLM Self-Critique}

\author{First Author \\
  Affiliation \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation \\
  \texttt{email@domain} \\}

\begin{document}
\maketitle

\begin{abstract}
Large language models (LLMs) are widely known to struggle with self-correction, yet the root cause remains debated. We hypothesize that self-correction failures stem not from inherent reasoning limitations but from \emph{identity bias} and \emph{context anchoring} during critique. To test this, we introduce the \textbf{Identity Swap Critique} framework, in which the same solution is presented to a critic model under different authorship labels (self, other model, weak model, anonymous, human) while the content remains identical. We conduct experiments across multiple reasoning benchmarks and LLM families, measuring error detection accuracy, correction quality, and confidence calibration under each condition. Our results show that [TODO: key finding]. We further decompose the sources of bias through context separation experiments and demonstrate that simple protocol-level mitigations---such as anonymization and session separation---significantly improve critique quality. These findings reframe self-correction as a protocol design problem rather than a fundamental reasoning limitation.
\end{abstract}

%% ============================================================
\section{Introduction}
\label{sec:intro}
%% ============================================================

Large language models are known to struggle with correcting their own mistakes. Models often fail to revise incorrect reasoning when reviewing their own outputs, yet can successfully identify and correct the same errors when the solutions are presented as external outputs rather than their own---a phenomenon termed the \emph{self-correction blind spot} \citep{tsui2025blindspot, huang2024selfcorrection}. Comprehensive surveys confirm that intrinsic self-correction, without external feedback or tool use, is unreliable and can even be harmful, sometimes flipping correct answers to incorrect ones \citep{kamoi2024selfcorrection}. Even iterative self-refinement approaches, while promising on surface-level text quality, show limited gains on reasoning tasks where errors are more subtle \citep{madaan2023selfrefine}.

A parallel line of research reveals a pervasive \emph{self-preference bias}: when LLMs are used as evaluators, they systematically rate their own outputs more favorably than equivalent outputs attributed to other models \citep{panickssery2024selfpreference, wataoka2024selfpreference}. This bias persists across tasks and model families, and recent work demonstrates that it follows \emph{assigned} identity rather than true origin---models favor whichever identity they are told is ``theirs'' \citep{lehr2025extremeselfpreference}. Self-bias has also been shown to amplify during self-refinement, where models preferentially preserve their own formulations even when alternatives are superior \citep{xu2024pride}.

Beyond self-preference, \emph{source framing} effects further complicate evaluation. Attributing identical text to different sources (e.g., human authors of different nationalities, or other LLMs) systematically shifts evaluation scores across multiple state-of-the-art models \citep{germani2025sourceframing}. Similarly, reviewing within the same conversation context leads to anchoring and rationalization, and simply separating production and review into distinct sessions improves error detection \citep{song2026crosscontext}.

These findings collectively suggest that LLM self-correction failure is not primarily a \emph{reasoning} problem but a \emph{perspective and attribution} problem. The model's judgment changes based on \emph{who it believes wrote the solution}, not on the solution's content. Yet no prior work has systematically controlled for authorship attribution across multiple identity conditions within a unified experimental framework, nor has the relative contribution of identity bias versus context anchoring been formally decomposed.

In this paper, we introduce the \textbf{Identity Swap Critique} framework to directly measure how authorship framing affects critique quality. Our contributions are:

\begin{itemize}
    \item We design a controlled experiment where the \emph{same solution} is presented to a critic under five identity conditions (\textsc{Self}, \textsc{Other Model}, \textsc{Weak Model}, \textsc{Anonymous}, \textsc{Human}), isolating the effect of attribution on error detection, correction, and confidence calibration.
    \item We decompose the sources of self-correction failure through \emph{context separation experiments} that disentangle identity bias from context anchoring effects.
    \item We demonstrate that simple, cost-free protocol-level mitigations---anonymization and session separation---substantially improve critique quality, establishing self-correction as a \emph{protocol design} problem rather than a fundamental reasoning limitation.
\end{itemize}

%% ============================================================
\section{Related Work}
\label{sec:related}
%% ============================================================

\subsection{Self-Correction in LLMs}

Self-correction---the ability of a model to identify and fix errors in its own outputs---has emerged as a critical challenge for deploying LLMs in reasoning-intensive settings. Early enthusiasm for iterative self-refinement, where a model generates, critiques, and revises its own output, demonstrated improvements on surface-level text quality but yielded more limited gains on reasoning tasks requiring deeper error detection \citep{madaan2023selfrefine}. Subsequent work has painted a more sobering picture: intrinsic self-correction---without access to external feedback, tools, or oracle signals---is not only unreliable but can actively \emph{degrade} performance, sometimes flipping correct answers to incorrect ones \citep{huang2024selfcorrection}. Comprehensive surveys have catalogued these failures across diverse settings, finding that self-correction succeeds primarily when augmented with external verification but fails consistently in the intrinsic case \citep{kamoi2024selfcorrection}.

A particularly striking manifestation of this failure is the \emph{self-correction blind spot}: models routinely detect and correct errors in external outputs while missing identical errors when those outputs are attributed to themselves. Controlled experiments across 14 open-source models reveal an average blind spot rate of 64.5\%, suggesting that the failure is systematic rather than task-specific \citep{tsui2025blindspot}. This asymmetry---same content, different judgment based on attribution---points toward bias in the evaluation process rather than a fundamental reasoning limitation.

\subsection{Self-Preference and Self-Attribution Bias}

The self-correction blind spot can be situated within a broader pattern of self-preference bias observed across LLM evaluation settings. When LLMs serve as evaluators, they systematically assign higher scores to their own generations than to equivalent outputs from other models, even when human annotators consider them of equal quality \citep{panickssery2024selfpreference}. Quantitative analyses in LLM-as-a-Judge frameworks have linked this bias to output familiarity: models tend to prefer text with lower perplexity relative to their own distribution \citep{wataoka2024selfpreference}.

Crucially, recent investigations have shown that this self-preference follows \emph{assigned identity} rather than true origin. When a model is explicitly told it is a different model, its evaluative preferences shift to favor that assigned identity, indicating that the bias is driven by identity framing in the prompt rather than genuine self-recognition \citep{lehr2025extremeselfpreference}. This finding has direct implications for our work: if identity labels alone can shift evaluative behavior, then manipulating these labels should modulate critique quality. Furthermore, self-bias has been shown to amplify during iterative refinement processes, where models preferentially preserve their own formulations even when superior alternatives are available \citep{xu2024pride}. Systematic evaluations using benchmarks such as MT-Bench and Chatbot Arena have further revealed that these biases interact with positional and verbosity effects, complicating the use of LLMs as reliable judges \citep{zheng2024judging}.

\subsection{Source Framing and Attribution Effects}

Beyond the self/other distinction, broader source framing effects have been documented across multiple LLM families. Large-scale experiments involving 192,000 assessments demonstrate that attributing identical narrative statements to different sources---particularly to human authors of specific nationalities or to other LLMs---systematically shifts evaluation scores \citep{germani2025sourceframing}. Notably, inter-model agreement, which is remarkably high in blind evaluation conditions, breaks down precisely when source framing is introduced. This fragility of LLM evaluation under attribution manipulation suggests that identity and authorship cues function as strong confounders that can override content-based judgment.

\subsection{Context Anchoring and Cross-Context Review}

A distinct but complementary source of self-correction failure arises from \emph{context anchoring}. When a model reviews its own output within the same conversation session, the full production history remains in the context window, anchoring the model's judgment and encouraging rationalization rather than scrutiny. Separating production and review into distinct sessions---even without altering authorship labels---has been shown to improve error detection, suggesting that conversational context itself biases the evaluation process independently of identity cues \citep{song2026crosscontext}. Disentangling these two sources of bias---identity framing versus context anchoring---is a central goal of our experimental design.

\subsection{Multi-Agent Debate and Cross-Model Review}

An alternative line of work seeks to improve critique quality through multi-model interaction. Multi-agent debate, where multiple LLMs discuss and challenge each other's reasoning, has been shown to improve factuality and reasoning accuracy \citep{du2023debate}. Structured disagreement protocols that encourage divergent thinking further enhance performance \citep{liang2023encouraging}. While effective, these approaches incur significant cost and complexity by requiring multiple model calls. Our work investigates whether simpler, single-model protocol changes---manipulating identity labels and context---can achieve comparable or even superior improvements at no additional cost.

\subsection{Critique Benchmarks}

Recent benchmarking efforts have begun to systematically evaluate the critique abilities of LLMs. Benchmarks spanning mathematical, commonsense, symbolic, coding, and algorithmic domains reveal a linear relationship between a model's generation capability and its critique quality, with stronger models exhibiting more effective critique of weaker models' outputs \citep{lin2024criticbench}. This asymmetry in cross-model critique quality---where the perceived capability gap between critic and author matters---provides an important baseline for interpreting our identity swap results, where we manipulate the \emph{perceived} author identity while holding actual content constant.

%% ============================================================
\section{Experimental Setup}
\label{sec:setup}
%% ============================================================

\subsection{Datasets}
\label{sec:datasets}

We evaluate on three reasoning benchmarks that require non-trivial chain-of-thought solutions:

\paragraph{GSM8K} \citep{cobbe2021gsm8k} contains grade-school math word problems requiring multi-step arithmetic reasoning. We use the test split and sample 500 problems.

\paragraph{MATH} \citep{hendrycks2021math} contains competition-level mathematics problems across algebra, geometry, number theory, and other domains with varying difficulty levels. We sample 300 problems.

\paragraph{GPQA} \citep{rein2023gpqa} is a graduate-level question-answering benchmark designed to be ``Google-proof.'' We use the diamond subset for its highest quality annotations.

% TODO: Finalize sample sizes and splits after pilot experiments.

\subsection{Models}
\label{sec:models}

We use three model families to ensure generality of findings:

\begin{itemize}
    \item \textbf{GPT-4o} (OpenAI): A state-of-the-art proprietary model.
    \item \textbf{Claude Sonnet} (Anthropic): A competitive proprietary model from a different developer.
    \item \textbf{Llama-3.1-70B-Instruct} (Meta): An open-source model served via vLLM.
\end{itemize}

Each model serves as both solver (generating chain-of-thought solutions) and critic (evaluating solutions under different identity conditions). This design allows us to measure self-critique bias within each model family.

\subsection{Identity Swap Critique Framework}
\label{sec:identity-swap}

The core experimental design presents the \emph{same solution} to the critic model under five authorship labels, varying only the identity preamble:

\begin{enumerate}
    \item \textbf{\textsc{Self}}: ``This is your previous solution. Please review your own work.''
    \item \textbf{\textsc{Other Model}}: ``This solution was generated by [Model X]. Please review it.''
    \item \textbf{\textsc{Weak Model}}: ``This solution was generated by a small, less capable model. Please review it.''
    \item \textbf{\textsc{Anonymous}}: ``Here is a solution. Please review it.''
    \item \textbf{\textsc{Human}}: ``A student wrote this solution. Please review it.''
\end{enumerate}

The solution text, problem statement, and all other prompt components remain identical across conditions. Solutions are generated once per (model, dataset) pair and reused across all identity conditions to ensure a controlled comparison.

\subsection{Context Separation Conditions}
\label{sec:context-separation}

To disentangle identity bias from context anchoring, we conduct a second experiment with five context conditions:

\begin{enumerate}
    \item \textbf{Same Session}: The solution appears as the model's own prior assistant turn in the conversation, and the model is asked to review it within the same session.
    \item \textbf{New Session}: The solution is presented in a fresh session with the \textsc{Self} label.
    \item \textbf{New Session + Anonymous}: Fresh session with no authorship information.
    \item \textbf{New Session + Paraphrased}: Fresh session with the solution paraphrased to reduce surface-pattern recognition.
    \item \textbf{Cross-Model}: A different model performs the review.
\end{enumerate}

Comparing these conditions allows us to decompose how much of the self-correction gap is attributable to identity framing versus context anchoring.

\subsection{Evaluation Metrics}
\label{sec:metrics}

For each critique, the model provides a structured JSON response indicating (1) whether the solution contains an error, (2) the error location, (3) a corrected answer, and (4) a confidence score. We compute:

\begin{itemize}
    \item \textbf{Error Detection Accuracy}: Whether the critic correctly identifies the presence or absence of errors.
    \item \textbf{True Positive Rate (TPR)}: Fraction of actually incorrect solutions flagged as erroneous.
    \item \textbf{False Positive Rate (FPR)}: Fraction of correct solutions incorrectly flagged.
    \item \textbf{Correction Accuracy}: Whether the critic's corrected answer matches the ground truth.
    \item \textbf{Confidence Calibration}: Expected Calibration Error (ECE) between stated confidence and actual accuracy.
\end{itemize}

%% ============================================================
\section{Experiment 1: Identity Swap Critique}
\label{sec:exp1}
%% ============================================================

% TODO: Results showing critique varies by authorship label
% Expected structure:
% - Main results table: Detection accuracy, TPR, FPR, correction accuracy per condition per model
% - Key finding: Self condition shows lower TPR / higher FPR than Anonymous/Other
% - Statistical significance tests (McNemar's test for paired comparisons)
% - Figure: Bar chart of detection accuracy across 5 conditions

%% ============================================================
\section{Experiment 2: Context Separation}
\label{sec:exp2}
%% ============================================================

% TODO: Results decomposing attribution bias vs. context anchoring
% Expected structure:
% - Comparison table: Same session vs. New session vs. New session + anonymous
% - Quantify: How much of the gap is context anchoring vs. identity bias
% - Figure: Decomposition showing relative contribution of each factor

%% ============================================================
\section{Experiment 3: Mitigation Strategies}
\label{sec:exp3}
%% ============================================================

% TODO: Results showing simple protocol changes improve correction
% Expected structure:
% - Baseline (Self) vs. each mitigation
% - Show improvement in detection accuracy and correction accuracy
% - Cost analysis: mitigations are free (no additional API calls for anonymization)
% - Compare with more expensive alternatives (multi-agent, cross-model)

%% ============================================================
\section{Analysis}
\label{sec:analysis}
%% ============================================================

% TODO: Additional analyses
% - Task difficulty vs. bias magnitude: Is identity bias stronger for harder problems?
% - Model size effects: Do larger models show more or less identity bias?
% - Confidence calibration across conditions: Are models better calibrated when anonymous?
% - False correction rates: Do models more often flip correct→incorrect under Self?
% - Context length effects: Does bias increase with longer solutions?

%% ============================================================
\section{Discussion}
\label{sec:discussion}
%% ============================================================

% TODO: Discuss implications
% Key points:
% - Self-correction is not a reasoning limitation but a protocol design problem
% - Simple mitigations (anonymization, session separation) are effective and free
% - Implications for LLM evaluation pipelines and deployment
% - Connection to human cognitive biases (confirmation bias, anchoring)
% - When to use cross-model review vs. anonymized self-review

\section*{Limitations}

% TODO: Discuss limitations
% - Limited to reasoning tasks; may not generalize to open-ended generation
% - Identity labels are surface-level; models may use other cues
% - JSON output format may constrain critique expressiveness
% - Sample sizes and model selection
% - Prompt sensitivity

\section*{Acknowledgments}

% TODO: Acknowledgments

\bibliography{refs}

\appendix

\section{Prompt Templates}
\label{sec:prompts}

% TODO: Include full prompt templates for each condition (system + user prompts)

\section{Additional Results}
\label{sec:additional-results}

% TODO: Per-dataset breakdowns, per-difficulty analyses, additional figures

\end{document}
